---
title: "Practical Statistics Exercises"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

##### KEY TERMS FOR DATA TYPES

Continuous
  Data that can take on any value in an interval.
  Synonyms
    interval, float, numeric
    
Discrete
  Data that can take on only integer values, such as counts.
  Synonyms
    integer, count
    
Categorical
  Data that can take on only a specific set of values representing a set of possible categories.
  Synonyms
    enums, enumerated, factors, nominal, polychotomous
    
Binary
  A special case of categorical data with just two categories of values (0/1, true/false).
  Synonyms
    dichotomous, logical, indicator, boolean
    
Ordinal
  Categorical data that has an explicit ordering.
  Synonyms
    ordered factor

There are two basic types of structured data: numeric and categorical. 

Numeric data comes in two forms:
Continuous, such as wind speed or time duration, 
Discrete, such as the count of the occurrence of an event. 

Categorical data takes only a fixed set of values, such as a type of TV screen (plasma, LCD, LED,
etc.) or a state name (Alabama, Alaska, etc.). 
Binary data is an important special case of categorical data
that takes on only one of two values, such as 0/1, yes/no, or true/false. 
Another useful type of categorical data is ordinal data in which the categories are ordered; an example of this is a numerical rating (1, 2, 3,
4, or 5).


##### KEY TERMS FOR RECTANGULAR DATA

Data frame
  Rectangular data (like a spreadsheet) is the basic data structure for statistical and machine learning models.

Feature
  A column in the table is commonly referred to as a feature.
  Synonyms
    attribute, input, predictor, variable

Outcome
  Many data science projects involve predicting an outcome — often a yes/no outcome. 
  The features are sometimes used to predict the outcome in an experiment or study.
  Synonyms
    dependent variable, response, target, output

Records
  A row in the table is commonly referred to as a record.
  Synonyms
    case, example, instance, observation, pattern, sample
    
    
    
#### Estimates of Location
Variables with measured or count data might have thousands of distinct values. A basic step in exploring
your data is getting a “typical value” for each feature (variable): an estimate of where most of the data is
located (i.e., its central tendency).

##### KEY TERMS FOR ESTIMATES OF LOCATION
Mean
  The sum of all values divided by the number of values.
  Synonyms
    average

$$Mean = \overline{x} = \frac{\sum_{i=1}^n x_i}{n}$$

Weighted mean
  The sum of all values times a weight divided by the sum of the weights.
  Synonyms
    weighted average

Median
  The value such that one-half of the data lies above and below.
  Synonyms
    50th percentile

Weighted median
  The value such that one-half of the sum of the weights lies above and below the sorted data.

$$Weighted\,mean = \overline{x}_w = \frac{\sum_{i=1}^{n} w_i x_i}{\sum_{i}^{n}w_i}$$

Trimmed mean
  The average of all values after dropping a fixed number of extreme values.
  Synonyms
    truncated mean
$$Trimmed\,mean = \overline{x} = \frac{\sum_{i=p+1}^{n-p} x_i}{n-2p}$$

Robust
  Not sensitive to extreme values.
  Synonyms
    resistant

Outlier
  A data value that is very different from most of the data.
  Synonyms
    extreme value
    
```{r mean}

state <- read.csv(file="C:/Users/Emilio Blanco/Documents/Emilio/DS/datascience-scripts/psds_data/state.csv")
mean(state[["Population"]])

mean(state[["Population"]], trim=0.1)

median(state[["Population"]])

```

The mean is bigger than the trimmed mean, which is bigger than the median.
This is because the trimmed mean excludes the largest and smallest five states (trim=0.1 drops 10%
from each end). If we want to compute the average murder rate for the country, we need to use a weighted
mean or median to account for different populations in the states. Since base R doesn’t have a function for
weighted median, we need to install a package such as matrixStats:

```{r weighted mean}
weighted.mean(state[["Murder.Rate"]], w=state[["Population"]])

library("matrixStats")
weightedMedian(state[["Murder.Rate"]], w=state[["Population"]])

```

#### Estimates of Variability
Location is just one dimension in summarizing a feature. A second dimension, __variability__, also referred to
as __dispersion__, measures whether the data values are tightly clustered or spread out. At the heart of
statistics lies variability: measuring it, reducing it, distinguishing random from real variability, identifying
the various sources of real variability, and making decisions in the presence of it.

##### KEY TERMS FOR VARIABILITY METRICS
Deviations
  The difference between the observed values and the estimate of location.
  Synonyms
    errors, residuals

Variance
  The sum of squared deviations from the mean divided by n – 1 where n is the number of data values.
  Synonyms
    mean-squared-error

$$Variance =s^2= \frac{\sum (x-\overline{x})^2}{n-1}$$

Standard deviation
  The square root of the variance.
  Synonyms
    l2-norm, Euclidean norm

$$Standard \,deviation =s= \sqrt{Variance}$$

Mean absolute deviation
  The mean of the absolute value of the deviations from the mean.
  Synonyms
    l1-norm, Manhattan norm

$$Mean\,Absolute\,Deviation = \frac{\sum_{i=1}^n |x_i-\overline{x}|}{n}$$

Median absolute deviation from the median
  The median of the absolute value of the deviations from the median.
  
$$Median\,Absolute\,Deviation = Median (|x_1-m|,|x_2-m|, ..., |x_N-m|)$$  
where m is the median. Like the median, the MAD is not influenced by extreme values. It is also possible
to compute a trimmed standard deviation analogous to the trimmed mean

Range
  The difference between the largest and the smallest value in a data set.

Order statistics
  Metrics based on the data values sorted from smallest to biggest.
  Synonyms
    ranks

Percentile
  The value such that P percent of the values take on this value or less and (100–P) percent take on this value or more.
  Synonyms
    quantile

Interquartile range
  The difference between the 75th percentile and the 25th percentile.
  Synonyms
    IQR


####Estimates Based on Percentiles
A different approach to estimating dispersion is based on looking at the spread of the sorted data.
Statistics based on sorted (ranked) data are referred to as order statistics. The most basic measure is the
range: the difference between the largest and smallest number. The minimum and maximum values
themselves are useful to know, and helpful in identifying outliers, but the range is extremely sensitive to
outliers and not very useful as a general measure of dispersion in the data.

To avoid the sensitivity to outliers, we can look at the range of the data after dropping values from each
end. Formally, these types of estimates are based on differences between percentiles. In a data set, the Pth
percentile is a value such that at least P percent of the values take on this value or less and at least (100 –
P) percent of the values take on this value or more. For example, to find the 80th percentile, sort the data.
Then, starting with the smallest value, proceed 80 percent of the way to the largest value. Note that the
median is the same thing as the 50th percentile. The percentile is essentially the same as a quantile, with
quantiles indexed by fractions (so the .8 quantile is the same as the 80th percentile).

A common measurement of variability is the difference between the 25th percentile and the 75th
percentile, called the interquartile range (or IQR). Here is a simple example: 3,1,5,3,6,7,2,9. We sort
these to get 1,2,3,3,5,6,7,9. The 25th percentile is at 2.5, and the 75th percentile is at 6.5, so the
interquartile range is 6.5 – 2.5 = 4. Software can have slightly differing approaches that yield different
answers (see the following note); typically, these differences are smaller.
For very large data sets, calculating exact percentiles can be computationally very expensive since it
requires sorting all the data values. Machine learning and statistical software use special algorithms, such
as [Zhang-Wang-2007], to get an approximate percentile that can be calculated very quickly and is
guaranteed to have a certain accuracy.

